Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2019-12-27 analysis/paper/paper.Rmd
some updates about background

There is only one type of locally made pottery found at the upper component of Kiwulan. There is around 1,200 pots were reconstructed showing the common round body form with a wide variety of impressed geometric motifs. Most sherds are fired to an orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes. It seems that there is no significant relationship between color and phases. Petrographic analysis for thin sections was conducted at the Department of Geosciences, National Taiwan University. The results shows high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate. Those minerals are believed to be local materials of the Yilan Plain. The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%). The size of the particles ranges from 500 micron to 1300 micron (=1 mm). Overall, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz.

@@

@@

## Metric measurements

In this study, we examine 291 pots recovered from Kiwulan site. The layer from 1 to 6 could be divided into 3 time periods. Layer 5 and 6 represent pre-contact period, layer 4 represents contact period, and layer1, 2, and 3 represent post-contact period.  The amount of pots for each layer shows below. Although most pots are not complete, the thickness from rim, neck, to body can be measured. I have also measured the diameter of rim, neck, and body. For those pots that are incomplete, the diameter is measured by its curvature. Since the height is incomplete for most pots, this preliminary analysis focuses more on the possible change in thickness and diameter of pot for different parts, and their ratio over time.

@@

>>>>

More than 550 thousands of potsherds were excavated from the upper component at Kiwulan, and around 1,200 vessels were reconstructed. Those vessels show high consistency in shape, which is globular body with short neck and wide mouth. A wide variety of impressed geometric motifs applied all over the exterior body surface are observed. They were believed as cooking pots according to the evidence of chars commonly observed on both interior and exterior surface. Most pots are fired to an orange to brownish color with a fully oxidized core or a reduced core with oxidized fringes.

## Analyses for pottery: Metric measurements and petrography

We examine 291 pots recovered from Kiwulan site. The layer from 1 to 6 could be divided into 3 time periods. Layer 5 and 6 represent pre-contact period, layer 4 represents contact period, and layer1, 2, and 3 represent post-contact period.  The amount of pots for each layer shows below. Although most pots are not complete, the thickness from rim, neck, to body can be measured. I have also measured the diameter of rim, neck, and body. For those pots that are incomplete, the diameter is measured by its curvature. Since the height is incomplete for most pots, this preliminary analysis focuses more on the possible change in thickness and diameter of pot for different parts, and their ratio over time.

Petrographic analysis for thin sections was conducted at the Department of Geosciences, National Taiwan University. The results show high percentage of inclusions (15-50%), including argillite, quartz, feldspar, metasandstone, sandstone fragment, and slate. The most common type of sherds, also the dominant one, has a high percentage of inclusion (25-40%), which is composed mainly by argillite (15-40%), followed by metasandstone (1-10%), sandstone (1-6%), and quartz (1-5%). The size of the particles ranges from 500 micron to 1300 micron (=1 mm). Overall, this type presents a mixture of fine, rounded argillite with a small part of rounded metasandstone and rounded to sub-angular monocrystalline quartz. Those minerals are believed to be local materials of the Yilan Plain.

@@

## Samples

@@

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the pottery used in this study. We thank Dr. Wen-Shan Chen in the Department of Geosciences, National Taiwan University for his invaluable guidance for petrographic analysis in his lab.
